Associations between Chlorophyll \emph{a} and various Microcystin-LR Health Advisory Concentrations

### 2015-06-02 manuscript/manuscript.Rmd
draft abstract rendered

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->

>>>>

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->
Cyanobacteria harmful algal blooms (cHABs) are associated with a wide arrange of adverse health effects that stem mostly from the presence of cyanotoxins.  To help protect agains the impacts, several health advisory levels have been set for some of these toxins, in particular, one of the most common toxins, microcystin, has several advisory levels set for drinking water and recreational use and managing water bodies to meet those levels could have far reaching benefits.  However, measuring micorcystin can not currently be done *in situ* and requires samples be processed in a lab.  This time consuming and expensive.  It is possible to find reliable indicators that may be estimated quickly and *in situ* as a first defense agains high level of microcystin.  In particular, chlorophyll *a* has been shown to be postively associated with microcystin.  In this paper we use this assocation to provide estimates of chlorophyll *a* that if exceeded would be indiciative of a higher likelihood of exceeding select concentrations of microcystin.  Using the 2007 National Lakes Assessment and a conditional probability appoach that has been used in other water quality settings, we idenfify chlorophyll *a* concentrations that are more likely than not to be associated with an exceedance of a microcystin health advisory level.  We look at the recent US EPA standards for drinking water as well as the World Health Organization levels for drinking water and recerational use.  For microcystin concentrations of 0.3, 1, 1.6, 2. and 4 we find chlorophyll *a* concentrations of [R CODE HERE] respectively.  When managing for these various microsystin levels exceed these reproted chlorophyll *a* concentratoins should be a trigger for further testing and possibly managment action.

### 2015-06-02 manuscript/manuscript.tex
re rendered with asis on table
draft abstract rendered

@@

>>>>

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
arrange of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect agains the impacts, several health advisory
levels have been set for some of these toxins, in particular, one of the
most common toxins, microcystin, has several advisory levels set for
drinking water and recreational use and managing water bodies to meet
those levels could have far reaching benefits. However, measuring
micorcystin can not currently be done \emph{in situ} and requires
samples be processed in a lab. This time consuming and expensive. It is
possible to find reliable indicators that may be estimated quickly and
\emph{in situ} as a first defense agains high level of microcystin. In
particular, chlorophyll \emph{a} has been shown to be postively
associated with microcystin. In this paper we use this assocation to
provide estimates of chlorophyll \emph{a} that if exceeded would be
indiciative of a higher likelihood of exceeding select concentrations of
microcystin. Using the 2007 National Lakes Assessment and a conditional
probability appoach that has been used in other water quality settings,
we idenfify chlorophyll \emph{a} concentrations that are more likely
than not to be associated with an exceedance of a microcystin health
advisory level. We look at the recent US EPA standards for drinking
water as well as the World Health Organization levels for drinking water
and recerational use. For microcystin concentrations of 0.3, 1, 1.6, 2.
and 4 we find chlorophyll \emph{a} concentrations of {[}R CODE HERE{]}
respectively. When managing for these various microsystin levels exceed
these reproted chlorophyll \emph{a} concentratoins should be a trigger
for further testing and possibly managment action.

@@

\begin{longtable}[c]{@{}lll@{}}
\toprule
\begin{minipage}[b]{0.13\columnwidth}\raggedright\strut
Source
\strut\end{minipage} &
\begin{minipage}[b]{0.18\columnwidth}\raggedright\strut
Microcystin
\strut\end{minipage} &
\begin{minipage}[b]{0.18\columnwidth}\raggedright\strut
Chlorophyll
\strut\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
EPA\_Child
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
0.3
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
23.68
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
1
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
65.2
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
EPA\_Adult
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
1.6
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
79.8
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
2
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
113.1
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
4
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
273.6
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
10
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
338.4
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.13\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
20
\strut\end{minipage} &
\begin{minipage}[t]{0.18\columnwidth}\raggedright\strut
338.4
\strut\end{minipage}\tabularnewline
\bottomrule
\end{longtable}
